Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-07-03 paper/manuscript.tex
Response to refs

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[11]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}

@@

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This happens typically in buried binding sites. Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}. Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.

\paragraph{Membrane proteins require additional care}
Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations. Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}). Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}. These systems require more extensive equilibration than soluble proteins. Some workflows, such as CHARMM-GUI or Desmond membrane preparation, recommend a multi-protocol to relax membrane systems \cite{https://doi.org/10.1002/jcc.23702,DesmondManual}.

Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water. Several approaches are reviewed in \cite{wu2022correction-51c}. The authors recommend the use of a co-alchemical ion, specifically making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.

In the case of a lipid-facing binding site, there are three additional challenges. First, simulation time should be extended to ensure sufficient sampling of lipid configurational space. Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand. If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols. Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids. Second, considering the free energy cost of a ligand partitioning into the membrane may provide additional insights \cite{doi:10.1021/acs.jcim.1c01147}. Third, the free energy of binding for a small molecule may be strongly coupled to the lipid species. In some cases, the identity of the lipid may be known. But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What simulation protocol should I choose   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{\tocorange{What simulation protocol should I choose?}}
\section{\tocorange{Choice of simulation protocol and implementation of alchemical transformations}}
  \sbnote{Emilio suggested to make the title more general; this is my suggestion!}
\tocsectioncomment{Steering group: Possibly expand to include other methods group agrees on.}
\tocsectioncomment{@Stefan Boresch to restructure as per his suggestion to remove single/dual topology section and begin with a more fundamental discussion which might more naturally include all options e.g. ATM. @Stefan Boresch to mention best practices for dummy atoms? This might benefit from following @Emilio Gallicchio's changes to make the theory section more general.}
\tocsectioncomment{@Sheenam Khuttan / @Emilio Gallicchio to contribute section on ATM.}
\tocsectioncomment{@Vytas Gapsys / @Volunteer to add best practices for non-equilibrium calculations.}
\label{sec:simulation_protocol_choice}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative''\footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state and in that sense are relative, while neither computes an absolute free energy.} (see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration. In contrast, in relative binding free energy calculations, we compare the binding free energy of two related ligands to determine the potency difference.

@@

\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules. For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology. In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc. (see \sbnote{appropriate cross-reference!}). The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages. In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B. The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state. We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates. The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously. Thus, we have dual coordinates, but chemical topologies for each of the entities. In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use. While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.

\paragraph{Single coordinate approaches --- alchemical topologies}

When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules. %Often this is predetermined by the choice of simulation software.
As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}. The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}. The single and dual topology approaches as proposed by Pearlman correspond to the left and middle paths of Fig.~\ref{fig:fig_topology}.  %\sbnote{Fig.~3 needs to be updated for dual coordinate methods and corrected, some bonds for the dummy atoms have the wrong anchor atom!}
The common substructure between the two molecules is the benzene ring, though in practice the substructure may be selected to be larger depending on the mapping chosen, as we discuss below.

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon. The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'' (labelled Du in Fig.~\ref{fig:fig_topology}), which retain their bonded interactions but do not interact with the rest of the system.  Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details. In a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}. At intermediate states, atoms behave as 'mixed' (M), either interpolating between two physical atoms, e.g., in the example, the hydrogen becomes a carbon, or between dummy and physical atoms.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
    \caption{\textbf{Strategies for alchemical calculations: single coordinate single and dual topology vs.\ dual coordinate dual topology.} \textbf{Left}: A single coordinate single topology converts from one type of atom to another. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms. At intermediate states, 'mixed' atoms (M) are present, either interpolating between physical atoms, e.g., a hydrogen to a carbon, or between physical and dummy atoms.  \textbf{Middle}: The single coordinate dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, typically using soft-core potentials for this. Here, 'mixed' intermediate atoms (M) always interpolate between physical and dummy atoms. Thus, 'mixed' atoms are used in both single coordinate approaches (single and dual topology), but the way the transformation occurs and the end states differ.
      %Following the arrow along the left and right illustrate the differences.
\textbf{Right}: In the dual coordinate dual topology approach both end states are present as separate molecules simultaneously.  Initially, benzyl alcohol is not interacting with the environment, at intermediate states the environment sees a mixture of the interactions of the two solutes, and, finally, benzene is not interacting with the environment. The two solutes are kept on top of each other by suitable restraints, indicated by the black ``spring'', and  the two molecules never see each other.  Figure adapted from the one used in V1 of this manuscript; adapted and extended by Sara Tkaczyk.
      % \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}
    }
    \label{fig:fig_topology}
\end{figure}

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.

%\sbnote{This probably needs to be rewritten\ldots}
As far as we are aware, the distinction between (single coordinate) single and dual topology approaches has been made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional. For an in-depth analyis of differences resulting from the use of a  single vs.\ a dual topology approach (single coordinate in both cases) see Ref.~\cite{Boresch_2002}.

\paragraph{Dual coordinate approaches --- dual chemical topologies}

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts (see below). %\sbnote{See below; where do we place the dummy atom discussion?}.
Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. The idea is illustrated in the right branch of Fig.~\ref{fig:fig_topology}.  Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. %In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

%Other methodologies also use both end states simultaneously; examples are the Separated Topologies (SepTop) approach [Baumann2023] and the ATM method [Azimi2022]. In the former, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to. Ries et al. refer to such an approach as “separated dual topology”, whereas they label their approach as “linked dual topology”. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space.

%One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.

\paragraph{Summary, comparison of single and dual coordinate methods, orthogonal classifications:}

%Some additional terms have also been employed to talk about these different intermediate pathways . Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.

Unfortunately, during the past years the use of the terms single and dual topology has proliferated, often with quite different meaning from the original intentions. The above presentation makes clear that the primary difference between approaches is the use of separate coordinates for each end state (chemical topology) or the use of an alchemical chimera with a single set of coordinates (alchemical topology). We therefore suggest the following classification
\begin{itemize}
\item Single coordinate (SC) --- Alchemical topology
  \begin{itemize}
  \item Single topology (SCST)
    \item Dual topology (SCDT)
  \end{itemize}
  \item Dual coordinate --- Chemical topology (DC)
  \end{itemize}
For completeness, we list some additional terminology that has been used: Some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though in our view these calculations should be classified as single coordinate single topology \sbnote{Check whether this is correct!}. %this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
Ries et al.\ suggested terminology \cite{Ries_2022} in which single topology corresponds to SCST, hybrid topology to SCDT, and dual topology to DC.

The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology. In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \cite{jiang2019computing,Lee_2023} or restraints. In the latter case, the restraints may couple the two set of coordinates directly \cite{Axelsen_1998,Ries_2022}, or the restraints may be defined relative to a common reference frame \cite{rocklin2013separated,Baumann_2023}. Finally, in ATM the two copies of the coordinates are held separate \cite{Azimi_2022}.

>>>>

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}

@@

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This can happen in buried binding sites~\cite{laage2017water}. Overlaying multiple protein X-ray structures can identify conserved water molecules or those that are associated only with certain ligands in a series and are therefore important to include in calculations. In cases where water molecules are known to play an important role in ligand binding, for example if the presence of electron density suggests they are tightly bound (indicated by a high EDIA score~\cite{nittinger2015evidence,meyder2017estimating}), software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}. Other tools such as WaterMap~\cite{abel2008role,young2007motifs} or open source equivalents (SSTMap~\cite{haider2018solvation}, GIST~\cite{ramsey2016solvation}, and others) can be used to define the hydration states for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.

In the context of relative binding free energy calculations, determining the correct hydration pattern for the two endpoints may not be sufficient for an accurate and reproducible calculation. If the locations of the bound water molecules differs at the two endpoints - e.g. if a perturbation to the ligand displaces a bound water molecule - then the network needs to be able to adapt during the perturbation. Otherwise, it is likely that the estimated relative binding free energies will be dependent on the direction in which the perturbation is performed~\cite{ross2020enhancing}.

In absolute binding free energy calculations, a ligand is fully decoupled from a binding site cavity which may otherwise be hydrated in the \textit{apo} state. In this case, the sampling of water molecules must be sufficient to hydrate the binding pocket as the ligand is decoupled (or vice versa for the reverse). When the binding site is exposed and water can move freely between the site and bulk solvent, then sampling is often not an issue. However, in the case of more buried sites or sites which form complex water networks, enhanced water sampling may be required to ensure an accurate binding affinity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What simulation protocol should I choose   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{\tocorange{What simulation protocol should I choose?}}
\tocsectioncomment{Steering group: Possibly expand to include other methods group agrees on.}
\tocsectioncomment{@Stefan Boresch to restructure as per his suggestion to remove single/dual topology section and begin with a more fundamental discussion which might more naturally include all options e.g. ATM. @Stefan Boresch to mention best practices for dummy atoms? This might benefit from following @Emilio Gallicchio's changes to make the theory section more general.}
\tocsectioncomment{@Sheenam Khuttan / @Emilio Gallicchio to contribute section on ATM.}
\tocsectioncomment{@Vytas Gapsys / @Volunteer to add best practices for non-equilibrium calculations.}
\label{sec:simulation_protocol_choice}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative''\footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state and in that sense are relative, while neither computes an absolute free energy.} (see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration. In contrast, in relative binding free energy calculations, we compare the binding free energy of two related ligands to determine the potency difference.

@@

\paragraph{Topologies} 
A critical first step is to determine which approach to use for atom mapping between end state molecules. Often this is predetermined by the choice of simulation software. Typically it is possible to chose between a \emph{dual topology} versus a \emph{single topology}

The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon. The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.  Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state. 
Thus, in a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology converts from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms.  \textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses soft-core potentials for this. The 'mixed' intermediate atoms (M) are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences. Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:fig_topology}
\end{figure}

As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.

Some additional terms have also been employed to talk about these different intermediate pathways . Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use. In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.

One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.

Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach. Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both). 
To our knowledge, efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.

\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed, meaning there is no MCSS match between the two compounds.}
    \label{fig:fig_mcss}
\end{figure}

There are different tools that allow the generation of MCSS matches as well as single topology input. A large number of software tools can compute MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}. The MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space. Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode. If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.

Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem. Using an absolute free energy approach instead seems more promising in such a case.

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
